A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation\tnoteref{ornl-cr}

### 2013-10-18 doc/jcomp2012/jcomp2012.tex
Last edits.

To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written as
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series}:
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation}
Thus, when $\rho(\vH) < 1$, we can recast the solution vector $\vx$ as
a series:
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
Using this series, we can write an iterative method, known as Richardson's
Iteration, that solves Eq.~(\ref{eq:Ax=b}):
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
Carlo methods that are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the adjoint
method to estimate the error in Eq.~(\ref{eq:MCSA-residual_solve}) instead of
the direct method outlined in \S~\ref{sec:direct-method}. To demonstrate the
effectiveness of the adjoint method over the direct method within the context
of MCSA, we choose the 2D time-dependent Poisson equation as a
simple model problem:
\begin{equation}
  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
  \label{eq:poisson_equation}
\end{equation}
For all comparisons, a single time step is computed with backwards Euler time
integration. The Laplacian is differenced on a square Cartesian grid with a
second-order five-point stencil,
\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}
and a fourth-order nine-point stencil,
\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}
both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
a single time step solution, we then have the following sparse linear system
to be solved with the MCSA method:
\begin{equation}
  \vA \vu^{n+1} = \vu^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}
Both the stencils will be used to vary the size and density of the sparse
linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).

>>>>

To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written as
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series}:
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation}
Thus, when $\rho(\vH) < 1$, we can recast the solution vector $\vx$ as
a series:
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
Using this series, we can write an iterative method, known as Richardson
iteration, that solves Eq.~(\ref{eq:Ax=b}):
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
Carlo methods that are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the adjoint
method to estimate the error in Eq.~(\ref{eq:MCSA-residual_solve}).
To demonstrate the
effectiveness of the adjoint method over the direct method within the context
of MCSA, we choose the 2D time-dependent Poisson equation as a
simple model problem:
\begin{equation}
  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
  \label{eq:poisson_equation}
\end{equation}
For all comparisons, a single time step is computed with backwards Euler time
integration. The Laplacian is differenced on a square Cartesian grid with a
second-order five-point stencil,
\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}
and a fourth-order nine-point stencil,
\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}
both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
a single time step solution, we then have the following sparse linear system
to be solved with the MCSA method:
\begin{equation}
  \vA \vu^{n+1} = \vu^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}
Both the stencils will be used to vary the size and density of the sparse
linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).
